All just Greedy for Money? - A prediction and Analysis of the interplay between Graduate Statistics and Salary Prospects in the STEM Field

### 2024-01-29 doc/DataLiteracyStudentProject/paper.tex
small changes

The analysis relies on three underlying datasets.
The first dataset captures the historical enrollment figures of students at the University of Tübingen. 
It contains the total number of students for each semester, for all study programs. This dataset is maintained and published by the
\href{https://uni-tuebingen.de/einrichtungen/verwaltung/iv-studierende/studierendenabteilung/statistiken/}{Statistics Bureau of the University of Tübingen}.
The University of Tübingen makes a good candidate to discover potential structure in the number
of enrollments because STEM programs are strongly represented here. % TODO: I think mentioning the potential risk of BIAS could be in the Discussion-part
To quantify the salary expectations, a dataset from the \href{https://www-genesis.destatis.de/genesis//online?operation=table&code=62321-0001&bypass=true&levelindex=0&levelid=1702307320529#abreadcrumb}{Deutsches Statistisches Bundesamt} is used.
It contains the salary trend for all major industrial sectors in Germany. This dataset does not adjust for inflation,
therefore an additional \href{https://www-genesis.destatis.de/genesis//online?operation=table&code=61121-0001&bypass=true&levelindex=0&levelid=1706446467491#abreadcrumb}{dataset} from the Deutsches Statistisches Bundesamt is used to account for that. It contains the inflation rate in Germany for each year.
The number of students is also limited by a further parameter. Namely the number of people eligible for an academic career.
The \href{https://www.statistik-bw.de/BildungKultur/SchulenAllgem/LRt0302.jsp}{Statistisches Landesamt Baden-Württemberg} provides
data about high school graduates in Baden-Württemberg for every year. We will not take other states into account, under
the assumption that the number of students leaving Baden-Württemberg, to study elsewhere, cancels out with others moving to the University of Tübingen.

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data. Combined with a
Partial autocorrelation function (PACF) the optimal value for the lag
is determined, see \Cref{sec:methods:lag}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{sec:methods} \textcolor{red}{subsection}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

\begin{figure*}
    \includegraphics{fig/prediction_5.pdf}
    \caption{VAR model Prediction with optimal lag value of $5$ on the test data set.}
\end{figure*}

@@

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.257\% & 1.271\% & 3.944\% \\
        \hline
        Prediciton 2 & 4.843\% & 0.119\% & 1.722\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to 1.}
\end{table}
The percentage differences between the true values especially for the students are quite high. To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$ \textcolor{red}{is $p$ = lag?}. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
The optimal lag based on AIC is 6, while for BIC it is 1, which is consistent with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Percentage difference between the true value and predicted value for students, salary, and graduates. Lag set to $5$.}
\end{table}

@@

With the correct lag chosen, the model fits the data by minimizing the mean square error.
The returned coefficients describe the impact of a specific lagged value from one time series
on predicting the next value. \ref{fig:student coefficients} visualizes all coefficients
relevant for predicting the number of students in the next semester. 
Notably, salary is the most relevant information for the prediction. All five
coefficient values for each time step are non-zero, indicating a strong correlation.
Graduates, on the other hand, are not as relevant, recognizable by values close to zero.

@@

To further understand the data underlying correlations, one can look at the correlation matrix
of residuals in table \ref{table:correlation matrix}.
This matrix describes the degree to which the model's errors, called residuals, are related to each other.
High correlations between errors can indicate a not fully captured underlying structure.
Values close to zero, like $-0.115$ between salary and graduates, indicate that the errors
in predicting the salary variable are not related to the errors in predicting the graduates variable.

@@

The training of the model stays set, and the whole dataset available is used. It is now possible to 
generate a new data set with a 10\% increase in salary expectations and set it as the new model input, figure \ref{fig:altered salary}.
The observed change in the prediction supports the obtained results in \Cref{sec:methods:fit}. There is a significant
dependence between salary expectations and the number of students enrolled in STEM courses. As soon as the salary
increases, there is also an upward trend visible in student numbers. The observed change is not massive which can be explained
by the fact that the mass of students are students in higher semesters. The only direct major influence on
the total number of enrolled students is made by people deciding to start this year. There won't be a lot of people changing subjects
halfway through their academic careers.

>>>>

The analysis is built on three underlying datasets.
The first dataset captures the historical enrollment figures of students at the University of Tübingen. 
It contains the total number of students for each semester, for all study programs. This dataset is maintained and published by the
\href{https://uni-tuebingen.de/einrichtungen/verwaltung/iv-studierende/studierendenabteilung/statistiken/}{Statistics Bureau of the University of Tübingen}.
The University of Tübingen makes a good candidate to discover potential structure in the number
of enrollments because STEM programs are strongly represented here. % TODO: I think mentioning the potential risk of BIAS could be in the Discussion-part
To quantify the salary expectations, a dataset from the \href{https://www-genesis.destatis.de/genesis//online?operation=table&code=62321-0001&bypass=true&levelindex=0&levelid=1702307320529#abreadcrumb}{Deutsches Statistisches Bundesamt} is used.
It contains the salary trend for all major industrial sectors in Germany. This dataset does not adjust for inflation,
therefore an additional \href{https://www-genesis.destatis.de/genesis//online?operation=table&code=61121-0001&bypass=true&levelindex=0&levelid=1706446467491#abreadcrumb}{dataset} from the Deutsches Statistisches Bundesamt is used to account for that. It contains the inflation rate in Germany for each year.
The number of students is also limited by a further parameter. Namely the number of people eligible for an academic career.
The \href{https://www.statistik-bw.de/BildungKultur/SchulenAllgem/LRt0302.jsp}{Statistisches Landesamt Baden-Württemberg} provides
data about high school graduates in Baden-Württemberg for every year. We will not take other states into account, under
the assumption that the number of students leaving Baden-Württemberg, to study elsewhere, cancels out with others moving to the University of Tübingen.

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data. Combined with a
partial autocorrelation function (PACF) the optimal value for the lag
is determined, see \Cref{sec:methods:lag}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{fig:accuracy model} \textcolor{red}{subsection}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

\begin{figure*}
    \includegraphics{fig/prediction_5.pdf}
    \caption{VAR model Prediction with optimal lag value of $5$ on the test data set.}
    \label{fig:accuracy model}
\end{figure*}

@@

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.257\% & 1.271\% & 3.944\% \\
        \hline
        Prediciton 2 & 4.843\% & 0.119\% & 1.722\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $1$.}
\end{table}
The percentage differences between the true values especially for the students are quite high. To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$ \textcolor{red}{is $p$ = lag?}. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
The optimal lag based on AIC is 6, while for BIC it is 1, which is consistent with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $5$.}
\end{table}

@@

With the correct lag chosen, the model fits the data by minimizing the mean square error.
The returned coefficients describe the impact of a specific lagged value from one time series
on predicting the next value. \Cref{fig:student coefficients} visualizes all coefficients
relevant for predicting the number of students in the next semester. 
Notably, salary is the most relevant information for the prediction. All five
coefficient values for each time step are non-zero, indicating a strong correlation.
Graduates, on the other hand, are not as relevant, recognizable by values close to zero.

@@

To further understand the data underlying correlations, one can look at the correlation matrix
of residuals in \Cref{table:correlation matrix}.
This matrix describes the degree to which the model's errors, called residuals, are related to each other.
High correlations between errors can indicate a not fully captured underlying structure.
Values close to zero, like $-0.115$ between salary and graduates, indicate that the errors
in predicting the salary variable are not related to the errors in predicting the graduates variable.

@@

The training of the model stays set, and the whole dataset available is used. It is now possible to 
generate a new data set with a 10\% increase in salary expectations and set it as the new model input, \Cref{fig:altered salary}.
The observed change in the prediction supports the obtained results in \Cref{sec:methods:fit}. There is a significant
dependence between salary expectations and the number of students enrolled in STEM courses. As soon as the salary
increases, there is also an upward trend visible in student numbers. The observed change is not massive which can be explained
by the fact that the mass of students are students in higher semesters. The only direct major influence on
the total number of enrolled students is made by people deciding to start this year. There won't be a lot of people changing subjects
halfway through their academic careers.
